Transfinite Games for Countable Nondeterminism

### 2017-03-01 nondeterminism/nondeterminism.tex
Some more modifications.

@@

@@

\begin{definition}
  Let $A=(M_A,\lambda_A,\zeta_A,P_A)$ be a game.  A (nondeterministic) \emph{strategy} for $A$ is a pair $\sigma=(T_\sigma, D_\sigma)$, where
  \begin{itemize}
    \item $T_\sigma$ is the set of \emph{traces} of the strategy; i.e., the set of positions that can occur when player $P$ plays according to the strategy.  That is, $T_\sigma$ is a non-empty prefix-closed subset of $P_A$ that is closed under $O$-replies to $P$-positions: whenever $s\in T_\sigma$ is a $P$-position and $a\in M_A$ is an $O$-move such that $sa\in P_A$, then $sa\in T_\sigma$.  We call such a set $T_\sigma$ a \emph{trace set}.

      We will often identify $T_\sigma$ with $\sigma$, and write $s\in\sigma$ instead of $s\in T_\sigma$.

      We say that $T_\sigma$ is \emph{compact} if it satisfies the compactness criterion given above for games: if $s\in M_A^\omega$ is an infinite sequence such that $t\in T_\sigma$ for all finite prefixes $t\ppprefix s$, then $t\in T_\sigma$.

      We require $T_\sigma$ to satisfy the following criterion:
      \begin{description}
        \item[Local compactness] For any $s\in T_\sigma$, there exists some compact trace set $T_s$ such that $s\in T_s\subset T_\sigma$.
      \end{description}

    \item $D_\sigma\subset T_\sigma$ denotes the set of \emph{divergences} in $T_\sigma$; that is, the set of those $O$-positions in $T_\sigma$ for which Player $P$ may elect to diverge rather than respond.  Accordingly, $\zeta_A(s)=O$ for all $s\in D_\sigma$.  
  \end{itemize}

  $T_\sigma$ and $D_\sigma$ are required to satisfy the following divergence axiom:
  \begin{description}
    \item[Diverge-or-reply] If $s\in T_\sigma$ is an $O$-position then either $sa\in T_\sigma$ for some $P$-move $a$ or there exists $t\prefix s$ such that $t\in D_\sigma$.  
  \end{description}

@@

@@

Let $A,B$ be games.  If $s\in \seqs{M_A\cprd M_B}$ then we write $s\vert_A$ for the subsequence of $s$ consisting of all moves from $A$ and $s\vert_B$ for the subsequence of $s$ consisting of all moves from $B$.  We write $P_A\|P_B$ for the set of all such sequences $s$ such that $s\vert_A\in P_A$ and $s\vert_B\in P_B$.

@@

To complete the proof that $T_{\comp\tau\sigma}$ is a legal trace set for a strategy, we need to show that $T_{\comp\tau\sigma}$ is closed under opponent replies and that it is locally compact.

@@

\begin{proposition}\label{locallyCompactComposition}
  Let $A,B,C$ be games, let $\sigma$ be a strategy for $A\implies B$ and let $\tau$ be a strategy for $B\implies C$.  Then $\comp\tau\sigma$ is locally compact.
\end{proposition}

Note that the limit of compact nondeterministic strategies need not be compact.

\begin{proposition}
  Let $A,B,C$ be games and let $\sigma\from A\implies B$ and $\tau\from B\implies C$ be (nondeterministic, locally compact) strategies.  Then $\comp\tau\sigma$ is locally compact.
  \begin{proof}
    Let $s\in T_{\comp\tau\sigma}$.  Then there is some sequence $\s\in\sigma\|\tau$ such that $\s\vert_{A,C}=s$.  We have $\s\vert_{A,B}\in\sigma$ and $\s\vert_{B,C}\in\tau$, so, by local compactness for $\sigma$ and $\tau$, there are compact strategies $\sigma'\from A\implies B$ and $\tau'\from B\implies C$ such that $\s\vert_{A,B}\in\sigma'$ and $\s\vert_{B,C}\in\tau'$.  After removing positions from $\sigma'$ and $\tau'$, we may assume that they are deterministic (and still compact).  Then $\comp{\tau'}{\sigma'}$ is compact and we have $\s\vert_{A,C}\in\comp{\tau'}{\sigma'}\subset\comp\tau\sigma$.
  \end{proof}
\end{proposition}

@@

\begin{definition}
  Let $A,B,C$ be games, let $\sigma$ be a strategy for $A\implies B$ and let $\tau$ be a strategy for $B\implies C$.  We define the divergences of $\comp\tau\sigma$ as follows:
  \[
    D_{\comp\tau\sigma} = \{\s\vert_{A,C}\suchthat\s\in\sigma\|\tau\textrm{ and either $\s\vert_{A,B}\in D_\sigma$ or $\s\vert_{B,C}\in D_\tau$}\}
    \]
\end{definition}

@@

    Let $\s'\in\sigma'\|\tau'$ be maximal such that $\s'\vert_{A,C}=s$.  We know that such a maximal sequence must exist: if $\s_i$ are a chain of sequences from $\sigma'\|\tau'$ such that $\s_i\vert_{A,C}=s$ then their limit $\s$ (which is contained in $\sigma'\|\tau'$ by compactness of $\sigma'$ and $\tau'$) must satisfy $\s\vert_{A,C}=\s$.  Since $s\not\in D_{\comp\tau\sigma}$, we must have $\s\vert_{A,B}\not\in D_\sigma$ and $\s\vert_{B,C}\not\in D_\tau$.

    Since $s$ is an $O$-position, one of $\s\vert_{A,B}$ and $\s\vert_{B,C}$ must be an $O$-position (by transitivity of $\Rightarrow$).  Suppose without loss of generality that $\s\vert_{A,B}$ is an $O$-position.  Since $\s\vert_{A,B}\not\in D_\sigma$, there must be some move $x$ such that $\s\vert_{A,B}x\in\sigma$, since $\sigma$ satisfies the diverge-or-reply rule.  Then we have $\s x\vert_{A,B}\in\sigma$.  Moreover, Lemma \ref{signProfilesLemma} tells us that $\s\vert_{B,C}$ must be a $P$-position, so $\s x\vert_{B,C}\in\tau$, since $\tau$ is closed under opponent replies.  Therefore, $\s x\in\sigma\|\tau$.  By maximality of $\s$, $x$ must be a move in $A$ and therefore we have $sx\in\comp\sigma\tau$.
  \end{proof}
\end{proposition}

@@

We make the category $\G$ into a symmetric monoidal closed category with tensor product given by $\tensor$ and linear implication given by $\implies$.  At the level of morphisms, the action of $\blank\tensor\blank$ is defined as follows: if $A,B,C,D$ are games, $\sigma$ is a strategy for $A\implies C$ and $\tau$ is a strategy for $B\implies D$, then $\sigma\tensor\tau$ is the strategy for $(A\tensor B)\implies (C\tensor D)$ given by playing the strategy $\sigma$ in $A$ and $C$ and playing the strategy $\tau$ in $B$ and $D$:
\[
  \sigma\tensor\tau = \left\{s\in P_{(A\tensor B)\implies (C\tensor D)}\suchthat s\vert_{A,C}\in\sigma,\;s\vert_{B,C}\in\tau\right\}
  \]
The divergences for $\sigma\tensor\tau$ are defined using the divergences for $\sigma$ and $\tau$:
\[
  D_{\sigma\tensor\tau} = \left\{s\in P_{(A\tensor B)\implies (C\tensor D)}\suchthat s\vert_{A,C}\in D_\sigma,\;s\vert_{B,D}\in D_\tau\right\}
  \]
\begin{proposition}
  $\sigma\tensor\tau$ is a valid strategy for $(A\tensor B)\implies (C\tensor D)$
  \begin{proof}
    $\sigma\tensor\tau$ is non-empty and prefix-closed, since $\sigma$ and $\tau$ are.  To show that it is a valid trace-set, we need to show that it is closed under opponent replies.

@@

The exponential connective gives rise to a functor $\G\to\G$, where the operation on morphisms $\sigma\from A\implies B$ is given by:
\[
  \oc\sigma = \{s\in P_{\oc A\implies \oc B}\suchthat s\vert_{A_n,B_n}\in\sigma\textrm{ for all $n\in\omega$}\}
  \]
The proof that $\oc(\cdot)$ preserves composition is the same as for the tensor product.

@@

>>>>

\newcommand{\stle}{{\;\le_s\;}}
\newcommand{\exle}{\sqsubseteq}
\newcommand{\exlub}{\bigsqcup}
\newcommand{\dv}{{\text{\lightning}}}

@@

We shall write $P_A^P=\zeta_A\inv(\{P\})$ for the set of $P$-positions in $A$ and $P_A^O=\zeta\inv(\{O\})$ for the set of $O$-positions in $A$.

@@

A (nondeterministic) \emph{strategy} for a game $A$ will be given by a pair $\sigma=(T_\sigma, D_\sigma)$, where
\begin{itemize}
  \item $T_\sigma$ is the set of \emph{traces} of the strategy; i.e., the set of positions that can occur when player $P$ plays according to the strategy.  That is, $T_\sigma$ is a non-empty prefix-closed subset of $P_A$ that is closed under $O$-replies to $P$-positions: whenever $s\in T_\sigma$ is a $P$-position and $a\in M_A$ is an $O$-move such that $sa\in P_A$, then $sa\in T_\sigma$.  We call such a set $T_\sigma$ a \emph{trace set}.

  \item $D_\sigma\subset T_\sigma$ denotes the set of \emph{divergences} in $T_\sigma$; that is, the set of those $O$-positions in $T_\sigma$ for which Player $P$ may elect to diverge rather than respond.  Accordingly, $\zeta_A(s)=O$ for all $s\in D_\sigma$.  If a trace has a divergent prefix, then it is divergent, so we shall follow Roscoe and insist that $D_\sigma$ be postfix-closed in $P_A^O$ (\cite{RoscoeCspInfinite}, cited in \cite{mcCHFiniteND}).
\end{itemize}

\begin{definition}
  Let $A=(M_A,\lambda_A,\zeta_A,P_A)$ be a game.  Then a \emph{strategy} for $A$ is a pair $\sigma=(T_\sigma,D_\sigma)$ such that
  \begin{itemize}
    \item $T_\sigma\subset P_A$ is a prefix-closed subset such that whenever $s\in T_\sigma$ is a $P$-position and $sa\in P_A$ for some move $a$, then $sa\in P_A$.
    \item $D_\sigma\subset P_A^O$ is a postfix-closed subset.
  \end{itemize}

  $T_\sigma$ and $D_\sigma$ are required to satisfy the following totality axioms:
  \begin{description}
    \item[Diverge-or-reply] If $s\in T_\sigma$ is an $O$-position then either $sa\in T_\sigma$ for some $P$-move $a$ or $s\in D_\sigma$.  
  \end{description}

  The diverge-or-reply rule tells us that a function must either return a value or go into an infinite loop, and corresponds to the absence of an error constant in our language.

@@

We will need one extra condition on the definition of a strategy in order to make sure that the composition of two strategies satisfies the diverge-or-reply rule.

\begin{definition}
  We say a strategy $\sigma$ is \emph{compact} if it satisfies the compactness criterion given above for games: if $s\in M_A^\omega$ is an infinite sequence such that $t\in T_\sigma$ for all finite prefixes $t\ppprefix s$, then $t\in T_\sigma$.  
\end{definition}

Compactness is too strong a condition for our purposes: we shall later see that it is not preserved by composition.  We shall use the following local version instead:

\begin{definition}
  We say a strategy $\sigma$ is \emph{locally compact} if whenever $s\in T_\sigma$ there is some compact strategy $\sigma'$ such that $s\in T_{\sigma'}\subset T_\sigma$ and $D_\sigma\subset D_{\sigma'}$.
\end{definition}

From now on, we shall use the word \emph{strategy} to mean \emph{locally compact strategy}.

@@

Let $A,B$ be games.  If $s\in \seqs{(M_A\cprd M_B)}$ then we write $s\vert_A$ for the subsequence of $s$ consisting of all moves from $A$ and $s\vert_B$ for the subsequence of $s$ consisting of all moves from $B$.  We write $P_A\|P_B$ for the set of all such sequences $s$ such that $s\vert_A\in P_A$ and $s\vert_B\in P_B$.

@@

To complete the proof that $T_{\comp\tau\sigma}$ is a legal trace set for a strategy, we need to show that $T_{\comp\tau\sigma}$ is closed under opponent replies.

@@

@@

\begin{definition}
  Let $A,B,C$ be games, let $\sigma$ be a strategy for $A\implies B$ and let $\tau$ be a strategy for $B\implies C$.  We define the divergences of $\comp\tau\sigma$ as follows: first we define
  \[
    \sigma\dv\tau = \left\{\s\in\seqs{(M_A\cprd M_B\cprd M_C)}\;\middle|\; \mbox{\pbox{\textwidth}{\flushRight either $\s\vert_{A,B}\in D_\sigma$ and $s\vert_{B,C}\in T_\tau$ \\ or $\s\vert_{A,B}\in T_\sigma$ and $s\vert_{B,C}\in D_\tau$}}\right\}
    \]
  \[
    D_{\comp\tau\sigma} = \{\s\vert_{A,C}\suchthat\s\in\sigma\dv\tau\}
    \]
\end{definition}

We need to show that this composition satisfies the diverge-or-reply and local compactness rules.

Note that the composition of compact nondeterministic strategies need not be compact.

\begin{proposition}
  Let $A,B,C$ be games and let $\sigma\from A\implies B$ and $\tau\from B\implies C$ be (nondeterministic, locally compact) strategies.  Then $\comp\tau\sigma$ is locally compact.
  \begin{proof}
    Let $s\in T_{\comp\tau\sigma}$.  Then there is some sequence $\s\in\sigma\|\tau$ such that $\s\vert_{A,C}=s$.  We have $\s\vert_{A,B}\in\sigma$ and $\s\vert_{B,C}\in\tau$, so, by local compactness for $\sigma$ and $\tau$, there are compact strategies $\sigma'\from A\implies B$ and $\tau'\from B\implies C$ such that $\s\vert_{A,B}\in\sigma'$ and $\s\vert_{B,C}\in\tau'$.  After removing positions from (and possibly adding divergences to) $\sigma'$ and $\tau'$, we may assume that they are deterministic (and still compact).  Then $\comp{\tau'}{\sigma'}$ is compact and we have $\s\vert_{A,C}\in\comp{\tau'}{\sigma'}\subset\comp\tau\sigma$ and $D_{\comp\tau\sigma}\subset D_{\comp{\tau'}{\sigma'}}$.
  \end{proof}
\end{proposition}

@@

    Let $\s'\in\sigma'\|\tau'$ be maximal such that $\s'\vert_{A,C}=s$.  We know that such a maximal sequence must exist: if $\s_i$ are a chain of sequences from $\sigma'\|\tau'$ such that $\s_i\vert_{A,C}=s$ then their limit $\s$ (which is contained in $\sigma'\|\tau'$ by compactness of $\sigma'$ and $\tau'$) must satisfy $\s\vert_{A,C}=\s$.  Since $s\not\in D_{\comp\tau\sigma}$, we must have $\s\vert_{A,B}\not\in D_\sigma\subset D_{\sigma'}$ and $\s\vert_{B,C}\not\in D_\tau\subset D_{\tau'}$.

    Since $s$ is an $O$-position, one of $\s\vert_{A,B}$ and $\s\vert_{B,C}$ must be an $O$-position (by transitivity of $\Rightarrow$).  Suppose without loss of generality that $\s\vert_{A,B}$ is an $O$-position.  Since $\s\vert_{A,B}\not\in D_{\sigma'}$, there must be some move $x$ such that $\s\vert_{A,B}x\in\sigma'$, since $\sigma'$ satisfies the diverge-or-reply rule.  Then we have $\s x\vert_{A,B}\in\sigma'$.  Moreover, Lemma \ref{signProfilesLemma} tells us that $\s\vert_{B,C}$ must be a $P$-position, so $\s x\vert_{B,C}\in\tau'$, since $\tau'$ is closed under opponent replies.  Therefore, $\s x\in\sigma'\|\tau'$.  By maximality of $\s$, $x$ must be a move in $A$ and therefore we have $sx\in\comp{\tau'}{\sigma'}\subset\comp\tau\sigma$.
  \end{proof}
\end{proposition}

@@

We make the category $\G$ into a symmetric monoidal closed category with tensor product given by $\tensor$ and linear implication given by $\implies$.  At the level of morphisms, the action of $\blank\tensor\blank$ is defined as follows: if $A,B,C,D$ are games, $\sigma$ is a strategy for $A\implies C$ and $\tau$ is a strategy for $B\implies D$, then $\sigma\tensor\tau$ is the strategy for $(A\tensor B)\implies (C\tensor D)$ given by playing the strategy $\sigma$ in $A$ and $C$ and playing the strategy $\tau$ in $B$ and $D$:
\[
  \sigma\tensor\tau = \left\{s\in P_{(A\tensor B)\implies (C\tensor D)}\suchthat s\vert_{A,C}\in\sigma,\;s\vert_{B,C}\in\tau\right\}
  \]
The divergences for $\sigma\tensor\tau$ are defined using the divergences for $\sigma$ and $\tau$:
\[
  D_{\sigma\tensor\tau} = \left\{s\in P_{(A\tensor B)\implies (C\tensor D)}\suchthat s\vert_{A,C}\in D_\sigma\textrm{ or }s\vert_{B,D}\in D_\tau\right\}
  \]
\begin{proposition}
  $\sigma\tensor\tau$ is a valid strategy for $(A\tensor B)\implies (C\tensor D)$
  \begin{proof}
    $\sigma\tensor\tau$ is non-empty and prefix-closed, since $\sigma$ and $\tau$ are.  To show that it is a valid trace-set, we need to show that it is closed under opponent replies.

@@

The exponential connective gives rise to a functor $\G\to\G$, where the operation on morphisms $\sigma\from A\implies B$ is given by:
\[
  \oc\sigma = \{s\in P_{\oc A\implies \oc B}\suchthat s\vert_{A_n,B_n}\in\sigma\textrm{ for all $n\in\omega$}\}
  \]
with divergences
\[
  D_{\oc\sigma} = \{s \in \oc\sigma\suchthat s\vert_{A_n,B_n}\in D_\sigma\textrm{ for some $n\in\omega$}\}
  \]
The proof that $\oc(\cdot)$ preserves composition is the same as for the tensor product.

@@

As usual, we shall use fixed point theorems from domain theory to define the denotation of $\Y$.  The main ideas come from \cite{LairdOrdinalGames}.

\subsection{The stable order on strategies}

We start by defining an order on strategies.  This is the order given in \cite{mcCHFiniteND}.

\begin{definition}
  Let $A$ be a game and let $\sigma,\tau$ be strategies for $A$.  We say that $\sigma\stle\tau$ if:

  \begin{enumerate}[1)]
    \item $T_\sigma\subset T_\tau$
    \item $D_\tau\subset D_\sigma$
    \item $T_\tau\subset T_\sigma \cup D_\sigma$
  \end{enumerate}
\end{definition}

To show that this is an order, the only non-trivial part is to show that property (3) is transitive.  Suppose that $\sigma\stle\tau\stle\upsilon$ are strategies and that $s\in T_\upsilon\setminus T_\tau$.  If $s\in T_\tau$, then there exists $s\in D_\sigma$, since $\sigma\stle\tau$ and if $s\not\in T_\tau$, then $s\in D_\tau\subset D_\sigma$, since $\tau\stle\upsilon$.  Therefore, $\sigma\stle\upsilon$.

\begin{proposition}
  Let $A,B,C$ be games, let $\sigma,\sigma'$ be strategies for $A\implies B$ and let $\tau,\tau'$ be strategies for $B\implies C$.  Suppose that $\sigma\stle\sigma'$ and $\tau\stle\tau'$.  Then:
  \[
    \comp\tau\sigma\stle\comp{\tau'}{\sigma'}
    \]
  \begin{proof}
    Since $T_\sigma\subset T_{\sigma'}$ and $T_\tau\subset T_{\tau'}$, we have $\sigma\|\tau\subset\sigma'\|\tau'$ and therefore $T_{\comp\tau\sigma}\subset T_{\comp{\tau'}{\sigma'}}$.

    Now suppose that $e\in D_{\comp{\tau'}{\sigma'}}$.  We want to show that there is some $d\in D_{\comp\tau\sigma}$ such that $d\prefix e$.

    We know that there exists some $\e\in\sigma'\|\tau$ such that $\e\vert_{A,C}=e$ and either $\e\vert_{A,B}\in D_{\sigma'}$ or $\e\vert_{B,C}\in D_\tau$.  Suppose first that $\e\vert_{A,B}\in\sigma$, so $\e\in\sigma\|\tau$.  Then we have either $\e\vert_{A,B}\in D_{\sigma'}\subset D_\sigma$ or $\e\vert_{A,B}\in D_\tau$, so $\e\vert_{A,C}=e\in D_{\comp\tau\sigma}$.

    Suppose instead that $\e\vert_{A,B}\not\in\sigma$.  We have $\e\vert_{A,B}\in\sigma'$ and $\sigma\stle\sigma'$, so there exists $d\in D_\sigma$ such that $d\prefix\e\vert_{A,B}$.  Let $\d$ be a prefix of $\e$ such that $\d\vert_{A,B}=d$.  Then we have $\d\vert_{A,B}\in D_\sigma$ and $\d\vert_{B,C}\in\tau$, so $\d\vert{A,C}\in\comp\tau\sigma$ and $\d\vert_{A,C}\prefix e$.

    Lastly, suppose that $s\in T_{\comp\tau{\sigma'}}\setminus T_{\comp\tau\sigma}$.  Write $s=\s\vert_{A,C}$, where $\s\in\comp\tau{\sigma'}$.  Then $\s\vert_{A,B}\in\sigma'\setminus\sigma$, and therefore there exists $d\in D_\sigma$ such that $d\prefix\s\vert_{A,B}$.  Let $\d\prefix\s$ be a prefix such that $\d\vert_{A,B}=d$.  Then $\d\vert_{A,C}\in D_{\comp\tau\sigma}$ and $\d\vert_{A,C}\prefix s$.

    A similar argument works in the case that $\sigma=\sigma'$.  Therefore, the result follows by transitivity of $\stle$.
  \end{proof}
\end{proposition}

We want to get a similar result for composition in the category $\G^{\oc}$.  To do that it will suffice to prove the following result:

\begin{proposition}
  Let $A,B$ be games and let $\sigma,\tau$ be strategies for $A\implies B$ such that $\sigma\stle\tau$.  Then $\oc\sigma\stle\oc\tau$ as strategies for $\oc A\implies \oc B$.
  \begin{proof}
    If $s\in\oc\sigma$, then $s\vert_{A_n,B_n}\in\sigma$ for all $n$ and therefore $s\vert_{A_n,B_n}\in\sigma'$ for all $n$.  It follows that $\sigma\subset\tau$.

    If $e\in\oc{\tau}$, then, since $\sigma\stle\tau$, for all $n$ either $e\vert_{A_n,B_n}\in\sigma$ or there exists $d_n\in D_\sigma$ such that $d_n\prefix e\vert_{A_n,B_n}$.  If $e\in D_{\oc\tau}$, then $e\vert_{A_n,B_n}\in D_\tau$ for some $n$, so there is at least one $d_n$.  We may choose a prefix $e'$ of $e$ such that $e\vert_{A_n,B_n}\prefix d_n$ whenever $d_n$ is defined and $d_n=e\vert_{A_n,B_n}$ for at least one $n$ (if at least one $d_n$ is finite, then one of the $d_n$ finishes first in $e$, and we let $e'$ be a minimal prefix of $e$ that restricts to $d_n$; if all the $d_n$ are infinite, let $e'=e$).  Since $D_\sigma\subset\sigma$, we see that $e'\in T_{\oc \sigma}$, and since $e'\vert_{A_n,B_n}=d_n$, we have $e'\in D_{\oc\sigma}$.

    Lastly, If $s\in \oc\tau\setminus\oc\sigma$, then there exists some $n$ such that $s\vert_{A_n,B_n}\in\tau\setminus\sigma$.  Therefore, $d_n$ is defined for at least one $n$ and it follows that we may construct $e'\in D_{\oc\sigma}$ as before such that $e'\prefix s$.
  \end{proof}
\end{proposition}

Let $A$ be a game and let $F$ be a set of strategies for $A$.  We define a strategy $\exlub F$ for $F$ by setting
\begin{gather*}
  T_{\exlub F} = \bigcup_{\sigma\in F}T_\sigma \\
  D_{\exlub F} = \bigcap_{\sigma\in F}D_\sigma
\end{gather*}
